Physcraper: A Python package for continually updated phylogenetic trees using the Open Tree of Life

### 2021-04-29 sources/bmc_template_submission/bmc_article.tex
some more edits

A user is interested in phylogenetic relationships within the genus \textit{Ilex}. Commonly
known as ``hollies", the genus encompasses between 400 \cite{gottlieb2005molecular}
and 500 accepted living species \cite{plantlist}, and is
the only extant taxon within the family Aquifoliaceae, in the order Aquifoliales of flowering
plants \cite{chase2016update}.

An online literature review in June 2020 (Google scholar search for ``ilex phylogeny")
reveals that there are several published phylogenies showing relationships within
the hollies \cite{cuenoud2000molecular, manen2010history, setoguchi2000intersectional,
selbach2009new}, but only two have data openly available, the ``Gottlieb2005" study
\cite{gottlieb2005molecular} and the ``Yao2020" study \cite{yao2020phylogeny}.
Gottlieb2005 made their original published phylogeny and alignment openly available in TreeBASE
(study 1091 \cite{treebase1091}). The Gottlieb2005 phylogeny samples 41 species, is available from
OpenTree's Phylesystem (study pg\_2827 \cite{pg2827}),
and has been integrated into OpenTree's synthetic tree \cite{mrcaott68451ott89474}.
The Yao2020 \textit{Ilex} phylogeny is the most recent one for the genus \cite{yao2020phylogeny},
and it is only available in OpenTree's Phylesystem (study ot\_1984 \cite{ot1984}),
and in the DRYAD repository \cite{yao2020dryad}.
With 175 tips, the Yao2020 phylogeny \cite{yao2020phylogeny} is the best sampled phylogeny available for
the genus \textit{Ilex}.
In order to showcase Physcraper's performance, we used the Gottlieb2005 phylogeny
and alignment to update relationships in the genus \textit{Ilex},
with the expectation to use the Yao2020 phylogeny, the best sampled and most recent
one, as a ``gold" standard to compare and verify Physcraper's results.

@@

To test the accuracy of Physcraper we performed a verification test in which
we pruned 9 out of the 41 tips in the original Gottlieb2005 phylogeny \cite{gottlieb2005molecular},
corresponding to a 20\% trim, excluding the outgroups. We then performed a Physcraper
run to test if we would recover the pruned tips. Physcraper automatically dropped
from the alignment the taxa pruned from the phylogeny to perform the BLAST searches.
The Physcraper updated phylogeny successfully recovered only 6 out of 9 pruned tips.
Closer examination of results revealed that sequences for the 3 missing tips were correctly retrieved
with BLAST along with the 6 sequences belonging to the remaining pruned tips.
We then followed the GenBank accession numbers reported in \cite{gottlieb2005molecular}
belonging to the sequences of the 3 missing tips, \textit{Ilex warburgii} (accession
number reported: U92600/U92601; updated: AH007153.2), \textit{Ilex dimorphophylla}
(accession number reported: U92592/U92593; updated: AH007149.2),
and \textit{Ilex percoriacea} (accession number: AH007156.2). We note that these three sequences
contain a 100 bp long gap of unidentified nucleotides (Ns) that is completely absent
from the original alignment. This caused them to exceed Physcraper's default sequence
length cutoff of 120\%, being thus filtered and excluded from further analyses.
The missing ITS sequences do appear in Physcraper's output file ``seqlen\_mismatch.txt",
that includes the accession number, taxon name, and sequence length of all BLAST matches
that were filtered based on sequence length cutoffs set in the configuration file.

>>>>

A user is interested in phylogenetic relationships within the genus \textit{Ilex}. Commonly
known as ``hollies", the genus encompasses between 400 \cite{gottlieb2005molecular}
and 500 recognized living species \cite{plantlist}, and is
the only extant taxon within the family Aquifoliaceae, in the order Aquifoliales of flowering
plants \cite{chase2016update}.

An online literature review in June 2020 (Google scholar search for ``ilex phylogeny")
revealed that there are several published phylogenies showing relationships within
the hollies \cite{cuenoud2000molecular, manen2010history, setoguchi2000intersectional,
selbach2009new}, but only two have data openly available, the ``Gottlieb2005" study
\cite{gottlieb2005molecular} and the ``Yao2020" study \cite{yao2020phylogeny}.
Gottlieb2005 made their original published phylogeny and alignment openly available in TreeBASE
(study 1091 \cite{treebase1091}). The Gottlieb2005 phylogeny samples 41 species, is available from
OpenTree's Phylesystem (study pg\_2827 \cite{pg2827}),
and has been integrated into OpenTree's synthetic tree \cite{mrcaott68451ott89474}.
The Yao2020 \textit{Ilex} phylogeny is the most recent one for the genus \cite{yao2020phylogeny},
and it is only available in OpenTree's Phylesystem (study ot\_1984 \cite{ot1984}),
and in the DRYAD repository \cite{yao2020dryad}.
With 175 tips, the Yao2020 phylogeny \cite{yao2020phylogeny} is the best sampled phylogeny available for
the genus \textit{Ilex}.
In order to showcase Physcraper's performance, we chose the Gottlieb2005 phylogeny
and alignment as Physcraper inputs to update relationships in the genus \textit{Ilex}.
Currently being the best sampled and most recent phylogenetic hypothesis for the genus,
we used the Yao2020 phylogeny as a golden standard to compare our results and verify
Physcraper's performance.

@@

To test the accuracy of Physcraper we performed a verification test in which
we pruned 9 out of the 41 tips in the original Gottlieb2005 phylogeny \cite{gottlieb2005molecular},
corresponding to a 20\% trim, excluding the outgroups. We then performed a Physcraper
run to test if we would recover the pruned tips. To perform the BLAST searches,
Physcraper automatically drops sequences from the alignment belonging to taxa
absent from the input phylogeny.
The Physcraper updated phylogeny successfully recovered only 6 out of 9 pruned tips.
Closer examination of results revealed that sequences for the 3 missing tips were correctly retrieved
with BLAST along with the 6 sequences belonging to the remaining pruned tips.
We then followed the GenBank accession numbers reported in \cite{gottlieb2005molecular}
belonging to the sequences of the 3 missing tips, \textit{Ilex warburgii} (accession
number reported: U92600/U92601; updated: AH007153.2), \textit{Ilex dimorphophylla}
(accession number reported: U92592/U92593; updated: AH007149.2),
and \textit{Ilex percoriacea} (accession number: AH007156.2). We note that these three sequences
contain a 100 bp long gap of unidentified nucleotides (Ns) that is completely absent
from the original alignment. This caused them to exceed Physcraper's default sequence
length cutoff of 120\%, being thus filtered and excluded from further analyses.
The missing ITS sequences do appear in Physcraper's output file ``seqlen\_mismatch.txt",
that includes the accession number, taxon name, and sequence length of all BLAST matches
that were filtered based on sequence length cutoffs set in the configuration file.
